---
title: "Chapter 3 — Probabilities and Hypothesis Testing (R companion)"
subtitle: "R/tidyverse translation of *Understanding Data by Alan Garfinkel and Yina Guo*"
author: "Adrian Soto-Mota, MD, PhD, FACP."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4, dpi = 120)
library(tidyverse)
set.seed(42)
```

# Orientation

Chapter 3 (book pp. 103–152) introduces **simulation-based NHST**: instead of invoking
a theoretical distribution, we construct the null world by computer and see how extreme
our observation is in that world. The book contains **no Python code listings** in this
chapter. This companion builds the machinery from loops (transparent, slow) up to
vectorized R (fast, idiomatic).

# §3.1 — Modeling and simulating probabilities  *(book p. 103)*

## A fair coin

```{r fair-coin}
# Simulate 1 toss: rbinom(n_experiments, size = n_tosses_per_experiment, prob)
rbinom(10, size = 1, prob = 0.5)          # 10 individual tosses, 0 = tails

# Simulate 1000 experiments of 10 tosses each, count heads in each
heads <- rbinom(1000, size = 10, prob = 0.5)
table(heads)
```

```{r coin-hist}
tibble(heads = heads) |>
  ggplot(aes(heads)) +
  geom_bar(fill = "steelblue") +
  scale_x_continuous(breaks = 0:10) +
  labs(title = "Distribution of heads in 1000 trials of 10 flips",
       x = "Number of heads", y = "Frequency")
```

## Rolling dice

```{r dice}
# Sample uniformly from {1..6}, with replacement, 10,000 times
rolls <- sample(1:6, size = 10000, replace = TRUE)
prop.table(table(rolls))   # empirical probabilities converge to 1/6
```

# §3.2 — Null Hypothesis Significance Testing by simulation  *(book p. 112)*

## The setup

Suppose we observe 7 heads out of 10 tosses of a coin. Is it "loaded"?

```{r nhst-setup}
observed_heads <- 7
n_tosses       <- 10
```

The null hypothesis: **the coin is fair**, `p = 0.5`.

## Simulate the null world

```{r simulate-null, fig.width=7, fig.height=4}
# Faithful loop version — this is what a pedagogical translation looks like
n_sim <- 10000
null_heads <- numeric(n_sim)
for (i in seq_len(n_sim)) {
  tosses <- sample(c("H", "T"), size = n_tosses, replace = TRUE)
  null_heads[i] <- sum(tosses == "H")
}

# Equivalent idiomatic R (much faster):
# null_heads <- rbinom(n_sim, size = n_tosses, prob = 0.5)

tibble(heads = null_heads) |>
  ggplot(aes(heads)) +
  geom_bar(fill = "grey70") +
  geom_vline(xintercept = observed_heads, color = "red", linewidth = 1) +
  scale_x_continuous(breaks = 0:10) +
  labs(title = "Null distribution with observed value (red)",
       subtitle = paste("Observed:", observed_heads, "heads out of", n_tosses),
       x = "Heads under the null", y = "Frequency")
```

## Compute the p-value

```{r compute-p}
# Two-sided p-value: "at least as extreme" in either direction.
# "As extreme" here means |heads − 5| >= |7 − 5| = 2
observed_distance <- abs(observed_heads - n_tosses/2)
p_value <- mean(abs(null_heads - n_tosses/2) >= observed_distance)
p_value
```

This matches what `binom.test()` gives us analytically:

```{r binom-test}
binom.test(x = observed_heads, n = n_tosses, p = 0.5)$p.value
```

## Surprise value  *(book §3.2, p. 117)*

The chapter introduces **surprise** `S = -log₂(p)` — the number of bits of
evidence against the null. It puts tiny p-values on a more intuitive scale:
`p = 0.05` → S ≈ 4.3 bits; `p = 0.001` → S ≈ 10 bits.

```{r surprise}
surprise <- function(p) -log2(p)
tibble(
  p  = c(0.5, 0.1, 0.05, 0.01, 0.001, 1e-6),
  S_bits = surprise(p)
)
```

# §3.2 — Factors that influence the p-value  *(book p. 125)*

Three knobs control how small a p-value you get:

1. **Effect size** — how far the true `p_coin` is from 0.5.
2. **Sample size** — `n_tosses`.
3. **Noise** — the inherent variability.

```{r knobs, fig.width=8, fig.height=4}
simulate_p <- function(n_tosses, true_p, n_sim = 5000) {
  null_heads     <- rbinom(n_sim, size = n_tosses, prob = 0.5)   # null world
  observed_heads <- rbinom(1,     size = n_tosses, prob = true_p) # one "real" expt
  obs_dist <- abs(observed_heads - n_tosses/2)
  mean(abs(null_heads - n_tosses/2) >= obs_dist)
}

grid <- expand_grid(
  n_tosses = c(10, 50, 200, 1000),
  true_p   = c(0.50, 0.55, 0.60, 0.70)
) |>
  rowwise() |>
  mutate(p_value = simulate_p(n_tosses, true_p)) |>
  ungroup()

ggplot(grid, aes(n_tosses, p_value, color = factor(true_p))) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  scale_x_log10() + scale_y_log10() +
  labs(title = "p-value as sample size and true effect vary",
       x = "n (log scale)", y = "p-value (log scale)",
       color = "true p")
```

Takeaway: with enough `n`, even a 0.55 coin (barely unfair) is flagged as
"significant". That's why the chapter insists on **reporting the effect size**,
not just the p-value.

# §3.3 — What's wrong with NHST?  *(book p. 132)*

Reproduce the chapter's critiques:

## (a) α = 0.05 is too lax  *(book §3.3, p. 132)*

```{r lax-alpha}
# If 100 studies per year test true nulls at α = 0.05,
# we expect about 5 "significant" false alarms every year.
# Across a whole literature, the false discovery rate is large.

# Across many experiments, the fraction of false positives at the chosen α is α.
n_exp <- 5000
alphas <- c(0.05, 0.01, 0.001)
p_values <- replicate(n_exp, {
  # True null: x ~ N(0,1), test H0: mean = 0
  x <- rnorm(30)
  t.test(x, mu = 0)$p.value
})
map_dbl(alphas, \(a) mean(p_values < a))
```

## (b) Confidence intervals are more informative  *(book §3.3, p. 142)*

```{r ci-better}
x <- rnorm(30, mean = 0.2, sd = 1)
t.test(x, mu = 0)$p.value         # Just a yes/no
t.test(x, mu = 0)$conf.int        # Tells us the plausible range of the true mean
```

A CI of, say, `[-0.05, 0.55]` is far more informative than "p = 0.07 (not
significant)".

## (c) p is not P(H0 | data)  *(book §3.3 "The p-value is the answer to the wrong question", p. 143)*

This is Chapter 12's full topic, but here's the preview:

```{r wrong-question}
# p-value answers: "If H0 were true, how often would we see data this extreme?"
# It does NOT answer: "Given these data, how likely is H0?"
# The second question requires Bayes' theorem and a prior probability.
```

# Exercises

1. Repeat the 7-out-of-10 analysis with observed = 70 out of 100. Is the p-value
   the same? Why not?
2. Flip the surprise formula. What p corresponds to "1 bit of evidence"? "20 bits"?
3. Using the `simulate_p()` function, find the smallest `n_tosses` at which a true
   `p_coin = 0.55` gives a p-value below 0.01 about 80 % of the time. (This is a
   power calculation, previewed for Chapter 11.)
