---
title: "Chapter 4 — Confidence Intervals (R companion)"
subtitle: "R/tidyverse translation of *Understanding Data*"
author: "Course notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4, dpi = 120)
library(tidyverse)
set.seed(42)
```

# Orientation

Chapter 4 (book pp. 153–186) defines the confidence interval by **resampling**
(the bootstrap), not by formula. In this way, every step is
transparent. The book contains **no Python code listings** in this chapter.
In here, we translate the chapter's "10,000 (virtual) friends" analogy (§4.2, p. 162) into
a for-loop first, then show the idiomatic one-liner.

# §4.1 — The idea  *(book p. 153)*

We have **one sample** `X` from some unknown population `U`. We'd love to know
the true mean μ, but we only see `X̄` (the sample mean).

**Insight:** if we had 10,000 sister samples, we could see how `X̄` varies
across them. We don't. But **bootstrapping treats `X` as a stand-in for `U`**
and resamples from `X` itself, with replacement.

# §4.2 — The bootstrap, written out  *(book p. 160)*

## Set up the observed sample

The chapter uses a biological example; we use an analogous one.

```{r one-sample}
# One observed sample: 30 measurements
observed_X <- c(12.1, 10.8, 11.5,  9.9, 11.0, 10.6, 12.3, 11.7, 10.2, 11.1,
                10.9, 11.8, 10.5, 11.2, 12.0, 10.7, 11.4, 11.6,  9.8, 10.4,
                11.3, 10.0, 11.9, 12.2, 10.3,  9.7, 11.0, 10.8, 11.5, 10.9)
n <- length(observed_X)
obs_mean <- mean(observed_X)
obs_mean
```

## The faithful loop

```{r bootstrap-loop}
n_resamples <- 10000
boot_means  <- numeric(n_resamples)

for (i in seq_len(n_resamples)) {
  # "Put 10,000 virtual friends back into X" — sample with replacement
  resample_i       <- sample(observed_X, size = n, replace = TRUE)
  boot_means[i]    <- mean(resample_i)
}

# Same thing, idiomatic R (one line):
# boot_means <- replicate(n_resamples, mean(sample(observed_X, n, replace = TRUE)))
```

## Construct the 95 % confidence interval

```{r construct-ci}
ci_95 <- quantile(boot_means, probs = c(0.025, 0.975))
ci_95
```

So our 95 % CI for the population mean μ is roughly `[ci_95[1], ci_95[2]]`.

```{r plot-boot, fig.width=7, fig.height=4}
tibble(boot_mean = boot_means) |>
  ggplot(aes(boot_mean)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = obs_mean, color = "red", linewidth = 1) +
  geom_vline(xintercept = ci_95, color = "red",
             linetype = "dashed") +
  labs(title = "Bootstrap distribution of the sample mean",
       subtitle = "Solid: observed mean.  Dashed: 2.5th and 97.5th percentiles (95% CI)",
       x = "Bootstrap mean", y = "Frequency")
```

# §4.3 — Worked example  *(book p. 168)*

Compare the bootstrap CI with the textbook t-interval:

```{r compare-t}
t_interval <- t.test(observed_X)$conf.int
tibble(
  method = c("Bootstrap", "t-distribution"),
  lower  = c(ci_95[1], t_interval[1]),
  upper  = c(ci_95[2], t_interval[2])
)
```

They agree closely because this sample is roughly Normal.
For skewed or small samples they diverge — the bootstrap is more honest about
the asymmetry.

# §4.4 — Bigger sample, tighter CI  *(book p. 172)*

The CI width should shrink as √n.

```{r width-vs-n, fig.width=7, fig.height=4}
# Vary n and bootstrap the CI
pop <- rnorm(100000, mean = 11, sd = 1)  # pretend this is the population

ci_width_for_n <- function(n, B = 2000) {
  samp       <- sample(pop, n)
  boot_means <- replicate(B, mean(sample(samp, n, replace = TRUE)))
  diff(quantile(boot_means, c(0.025, 0.975)))
}

widths <- tibble(
  n       = c(10, 20, 50, 100, 200, 500, 1000),
  width   = map_dbl(n, ci_width_for_n)
)

ggplot(widths, aes(n, width)) +
  geom_line() + geom_point() +
  scale_x_log10() + scale_y_log10() +
  labs(title = "CI width shrinks like 1/√n",
       x = "Sample size n (log)", y = "95% CI width (log)")
```

Slope ≈ −0.5 on this log-log plot confirms the √n rule.

# §4.5 — Validating the resampling principle  *(book p. 177)*

Does a 95 % CI actually contain the true mean 95 % of the time? Let's check.

```{r coverage}
true_mu <- 11                     # known population mean
n        <- 30
n_trials <- 1000                  # repeat the whole experiment 1000 times
B        <- 500                   # bootstrap replicates per trial

captured <- logical(n_trials)
for (t in seq_len(n_trials)) {
  sample_t   <- rnorm(n, mean = true_mu, sd = 1)
  boot_means <- replicate(B, mean(sample(sample_t, n, replace = TRUE)))
  ci         <- quantile(boot_means, c(0.025, 0.975))
  captured[t] <- (ci[1] <= true_mu) && (true_mu <= ci[2])
}
mean(captured)  # should be ≈ 0.95
```

# §4.6–4.7 — Other definitions & formulas  *(book p. 180)*

For the record, the textbook Normal-based formula is:

```{r formula-ci}
X_bar <- mean(observed_X)
SE    <- sd(observed_X) / sqrt(length(observed_X))
t_critical <- qt(0.975, df = length(observed_X) - 1)

lower <- X_bar - t_critical * SE
upper <- X_bar + t_critical * SE
c(lower = lower, upper = upper)
```

This agrees with `t.test()$conf.int`. It requires the Normality assumption; the
bootstrap does not.

# §4.8–4.9 — How to present and interpret a CI  *(book pp. 181–182)*

**Report:** `mean = 11.02 (95% CI 10.74 to 11.30)`.

**Interpret, correctly:** "If we repeated this experiment many times, 95 % of
such intervals would contain the true population mean."

# Exercises

1. Replace `observed_X` with `rexp(30, rate = 0.1)` (exponential, heavily skewed).
   Compare the bootstrap CI to `t.test()$conf.int`. Which one gives an asymmetric
   interval? Why is that more honest?
2. Run the coverage simulation (§4.5) with `n = 5`. Is the bootstrap still ≈ 95 %?
   Why might it under-cover for very small samples?
3. Implement the **percentile bootstrap** (shown here) and the **basic
   bootstrap** `2·θ̂ − q₁₋α/₂, 2·θ̂ − q_{α/2}`. Compare.
