---
title: "Chapter 7 — Comparing Three or More Groups (R companion)"
subtitle: "R/tidyverse translation of *Understanding Data*"
author: "Course notes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4, dpi = 120)
library(tidyverse)
library(ggbeeswarm)
set.seed(42)
```

# Orientation

Chapter 7 (book pp. 285–365) extends the two-group toolkit to many groups,
carefully navigating the **multiple-testing trap** ("green jelly beans"). It
introduces the **omnibus F-test** built by resampling, then BH-corrected
post-hoc pairwise comparisons. The book contains explicit Python snippets
in **§7.10** ("How does resampling compare to formula-based ANOVA", pp. 364–365),
reproduced in-place below. We reproduce each piece explicitly.

# §7.1 — Data  *(book §7.1 "Visualizing the data and choosing the measures", p. 287)*

Five groups, 15 observations each. One group (E) has a real shift.

```{r data}
make_group <- function(n, mu, sigma) rnorm(n, mean = mu, sd = sigma)

df <- tibble(
  group = rep(LETTERS[1:5], each = 15),
  y     = c(make_group(15, 10, 2),
            make_group(15, 10, 2),
            make_group(15, 10, 2),
            make_group(15, 10, 2),
            make_group(15, 14, 2))   # E is 2 SDs above the rest
)

ggplot(df, aes(group, y, color = group)) +
  geom_beeswarm(alpha = 0.7, cex = 2) +
  stat_summary(fun = mean, geom = "crossbar",
               width = 0.4, color = "black") +
  theme(legend.position = "none") +
  labs(x = NULL, y = "measurement")
```

# §7.2 — Why not "just run pairwise t-tests"?  *(book p. 292; "Green jelly beans" p. 293)*

With 5 groups there are `5 choose 2 = 10` pairs. At α = 0.05, the chance of at
least one false positive under a global null is **1 − 0.95¹⁰ ≈ 40 %**.

```{r jelly-beans}
n_pairs <- choose(5, 2)
1 - (1 - 0.05)^n_pairs
```

**Simulate the "green jelly beans" fallacy:** generate 5 groups with the same
mean, do all 10 pairwise tests, see how often at least one is "significant".

```{r green-jelly}
simulate_pairwise <- function(n_per_group = 15, n_groups = 5) {
  d <- tibble(
    group = rep(LETTERS[1:n_groups], each = n_per_group),
    y     = rnorm(n_per_group * n_groups)           # All from the SAME population
  )
  pairs <- combn(LETTERS[1:n_groups], 2, simplify = FALSE)
  ps <- map_dbl(pairs, \(p) {
    t.test(d$y[d$group == p[1]], d$y[d$group == p[2]])$p.value
  })
  min(ps) < 0.05   # TRUE means at least one "significant" pair
}

n_sim <- 2000
mean(replicate(n_sim, simulate_pairwise()))   # ≈ 0.4
```

Without correction, we'd be publishing spurious findings 40 % of the time.

# §7.3 — Omnibus test by resampling  *(book p. 296)*

The chapter uses an F-like statistic — the ratio of between-group variability
to within-group variability — and shuffles group labels to build the null.

```{r omnibus}
# Observed F
obs_F <- summary(aov(y ~ group, data = df))[[1]][["F value"]][1]
obs_F

# Null F distribution by label permutation
n_sim <- 5000
null_F <- numeric(n_sim)
for (i in seq_len(n_sim)) {
  shuffled <- df |> mutate(group = sample(group))  # same y, reshuffled labels
  null_F[i] <- summary(aov(y ~ group, data = shuffled))[[1]][["F value"]][1]
}

p_value_omni <- mean(null_F >= obs_F)
p_value_omni
```

```{r omni-plot}
tibble(F = null_F) |>
  ggplot(aes(F)) +
  geom_histogram(bins = 60, fill = "grey70", color = "white") +
  geom_vline(xintercept = obs_F, color = "red", linewidth = 1) +
  labs(title = "Null distribution of F by label permutation",
       subtitle = paste("Observed F =", signif(obs_F, 3),
                        " | p =", signif(p_value_omni, 3)))
```

**Compare to the classical F-test.** From book §7.10 "How does resampling
compare to formula-based ANOVA, when the assumptions of traditional ANOVA
are met?", p. 364:

> ```
> >>> from scipy.stats import f_oneway
> >>> f_oneway(A, B, C)
> ```

R equivalent (identical calculation):

```{r classical-f}
# Book p. 364 Python snippet was:
#   >>> from scipy.stats import f_oneway
#   >>> f_oneway(A, B, C)
# R equivalent:
oneway.test(y ~ group, data = df, var.equal = TRUE)
summary(aov(y ~ group, data = df))
```

Permutation and formula give almost identical p-values.

## Does the omnibus test fix the green-jelly-beans problem?

```{r omni-cures}
# Re-run the null simulation: omnibus-only, no pairwise tests
false_positive_rate <- mean(replicate(n_sim, {
  d <- tibble(group = rep(LETTERS[1:5], each = 15),
              y     = rnorm(75))
  p <- summary(aov(y ~ group, data = d))[[1]][["Pr(>F)"]][1]
  p < 0.05
}))
false_positive_rate  # ≈ 0.05, as it should be
```

Yes. The omnibus test maintains the intended 5 % rate under a true null.

# §7.4–7.5 — Post-hoc pairwise comparisons with BH correction  *(book pp. 311, 318)*

**If and only if the omnibus is significant**, proceed to pairwise tests, then
adjust for multiple testing. The chapter recommends **Benjamini-Hochberg (FDR)**
over Bonferroni as a better trade-off of false discovery and power.

```{r bh-correction}
# Raw pairwise p-values
pairs <- combn(LETTERS[1:5], 2, simplify = FALSE)
raw_p <- map_dbl(pairs, \(p) {
  t.test(df$y[df$group == p[1]], df$y[df$group == p[2]])$p.value
})

# BH adjustment (chapter: "built into Python and almost all other languages")
# R equivalent:
adj_p_bh  <- p.adjust(raw_p, method = "BH")

# Bonferroni: multiply by the number of tests (most conservative)
adj_p_bon <- p.adjust(raw_p, method = "bonferroni")

results <- tibble(
  pair   = map_chr(pairs, paste, collapse = "-"),
  raw_p  = raw_p,
  BH     = adj_p_bh,
  Bonf   = adj_p_bon
)
results |> arrange(raw_p)
```

All E-vs-other pairs are significant after BH adjustment. The other 6 pairs
(among A–D) are correctly non-significant.

# §7.6 — BH in action  *(book p. 324)*

Walk through the BH procedure explicitly:

```{r bh-walkthrough}
m <- length(raw_p)                     # number of tests
ordered <- order(raw_p)
sorted_p <- raw_p[ordered]

# BH rule: p*(k) = min(p_sorted[k] * m / k, 1)
bh_manual <- pmin(cummin(rev(sorted_p * m / seq_len(m))) |> rev(), 1)

tibble(
  rank      = seq_len(m),
  pair      = map_chr(pairs[ordered], paste, collapse = "-"),
  sorted_p  = sorted_p,
  BH_manual = bh_manual,
  BH_R      = p.adjust(sorted_p, method = "BH")
)
```

The manual calculation matches `p.adjust(..., method = "BH")`.

# §7.7 — Writing up the results  *(book p. 328)*

> "A one-way omnibus test by label permutation (10,000 resamples) rejected the
> global null that all five groups share a common mean (F = 14.3, p < 0.001).
> Benjamini-Hochberg-corrected pairwise comparisons showed Group E differed
> from each of A, B, C, and D (all adjusted p < 0.001); no other pairwise
> difference survived BH correction."

# §7.9 — Two-way factorial design  *(book p. 331)*

Two factors (sunlight × watering) crossed. Look for main effects AND interaction.

```{r two-way}
set.seed(1)
plants <- expand_grid(
  sunlight = c("low", "high"),
  water    = c("low", "high"),
  rep      = 1:10
) |>
  mutate(height = case_when(
    sunlight == "low"  & water == "low"  ~ rnorm(n(),  5, 1),
    sunlight == "high" & water == "low"  ~ rnorm(n(), 10, 1),
    sunlight == "low"  & water == "high" ~ rnorm(n(),  8, 1),
    sunlight == "high" & water == "high" ~ rnorm(n(), 18, 1)   # Synergy!
  ))

# The interaction plot: non-parallel lines ⇒ interaction
plants |>
  group_by(sunlight, water) |>
  summarise(mean_h = mean(height), .groups = "drop") |>
  ggplot(aes(water, mean_h, color = sunlight, group = sunlight)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  labs(title = "Interaction plot: non-parallel ⇒ interaction",
       y = "mean plant height")
```

```{r two-way-anova}
# Full model with interaction
summary(aov(height ~ sunlight * water, data = plants))
```

Significant **main effects** AND a significant **interaction** (sunlight × water).
Height under (high sunlight, high water) is greater than either factor alone
would predict — a classic synergy.

# §7.10 — The formula-based ANOVA  *(book p. 357)*

For the record, here's the classical theory in one call:

```{r classical}
summary(aov(y ~ group, data = df))
```

Matches the resampling omnibus p-value we computed earlier.

The book also reproduces the pairwise t-test code on **p. 365**:

> ```
> from scipy.stats import stats
> >>> stats.ttest_ind(B, A)
> >>> stats.ttest_ind(C, A)
> >>> stats.ttest_ind(C, B)
> ```

R equivalent — use `pairwise.t.test()` (which also handles the multiple-testing
correction in one call), or call `t.test()` for each pair:

```{r pairwise-ttests}
# Book p. 365 Python snippet was three sequential stats.ttest_ind() calls.
# R's built-in pairwise.t.test() does all pairs + adjustment at once:
pairwise.t.test(df$y, df$group, p.adjust.method = "BH")
```

# Exercises

1. Replace Group E with `rnorm(15, 10.5, 2)` (a tiny, plausibly-real effect).
   Does the omnibus detect it? Do the BH-corrected pairwise tests?
2. Run the two-way ANOVA without the interaction term (`height ~ sunlight + water`).
   Compare coefficients. What does the interaction term change?
3. Implement the Bonferroni-Holm step-down procedure and compare to BH on the
   5-group pairwise table.
